### 2023-06-13 manuscript/chapters/3_mappings.qmd
nearly finished chap 3 !

## Population dynamics : an introductory example of a 1D embedding

@@

Let us for a moment consider another type of model networks of neurons, that of homoegeneous population dynamics, that describes how $N_\text{pop}$ different populations interact with each other. These types of models emerge when networks of neurons exhibit a finite number of "neuron types". We then group together similar neurons, and make the approximation that all neurons inside the population have the same potential, and the same interaction weights with other neurons. In effet, every population then represents in actuality an infinity of discrete identical neurons.

@@

We don't show the derivation of population dynamics here (see @GerKis14, chapter 12.3 for a detailed analysis), but we point out the implicit scaling behavior of $\tilde J_{ab}$. Letting $i$ be a neuron of population $a$ and $j$ be neuron of population $b$, the connectivity between the two neurons is $J_{ij}$ and the population connectivity scales with the number of neurons in the presynaptic population $|b|$. Furthermore, we make the reasonable assumption that $J_{ij}$ scales as $J^0_{ij} / N$, where $J^0_{ij} \sim \bO(1)$.

@@

So far, we have not considered that the populations are embedded in some abstract space ; but the notion of population weights gives rise to the obvious embedding in $[0,1]$ given by the cumulative sums of the population weights, illustrated in @fig-popdynamics-01. We define (somewhat informally) the connectivity kernel $w(\alpha, \beta) = J^0_{ab}$ for $\alpha, \beta$ in the regions corresponding to populations $a$ and $b$ respectively[^poplocation].

@@

When taken in isolation, the previous developments might sound a little bit silly, though they serve to give an impression of the problem at hand, and an interpretation of the rest of this work. For a finite number of populations $N_\text{pop}$, the population dynamics is perfectly defined, however what happens as $N_\text{pop} \to \infty$ ? The core of the question lies in the fact that we can easily *enumerate* a finite number of populations, but the sum $\sum_{b=1}^{N_\text{pop}}$ must have a clearly defined interpration when this number tends to infinity.

In the light of the toy model presented in @sec-nf, we can reinterpret the density $\rho$ of the neural field equation. Every point $\vec z \in \R^p$ (let us consider $p=2$ for the rest) describes a infinite population of homogeneous neurons, associated with a probability (density) $\rho(\vec z)$. Therefore, in order to write an equivalent neural field in $[0,1]$, we need a way to *enumerate* the populations. This is done using a mapping $S : [0,1]^2 \mapsto [0,1]$, where we traverse the $[0,1]^2$ space by following the order in the image $[0,1]$.

@@

For a finite number of populations $N_\text{pop}$, applying a bijective mapping $S$ will evidently give an equivalent neural field, because the connectivity matrix is permutation-invariant, and the resulting connectivity matrix $J_{S(a),S(b)}$ will therefore give equivalent dynamics. As we increase $N_\text{pop}$ by taking finer and finer grids, the image of the grids define a sequence of mappings $S^n : \text{grid in}\ [0,1]^2 \mapsto \text{grid in}\ [0,1]$, which we aim to study in the rest of this chapter.

@@

The connectivity kernel $\tilde w$ defines the interactions between the populations in the $[0,1]$ embedding, which we define in relation to the uniform kernel on $[0,1]^2$ as :

@@

When we do the change of variables with the CDF, we don't encounter problems because the gaussian (inverse) CDF is differentiable and its derivative is continuous[^inversecdf] ; therefore $w_U(\vec v, \vec u) = w(\textrm{CDF}^{-1}(\vec v), \textrm{CDF}^{-1}(\vec u))$ is still highly regular.

[^inversecdf]: Strictly speaking, we have the problem that $\textrm{CDF}^{-1}(v) \xrightarrow{v \to 1} \infty$, and so the derivative diverges when $v \to 1$ (and similarly, when $v \to 0$). However this is not a problem, since these limits correspond to $z \to \pm \infty$, where the density vanishes.

@@

We start our search of mappings by attempting to find a mapping $S$ that is differentiable.
The easiest example of such a mapping would be a linear projection, for instance $P(v_1, v_2) = v_1$. This mapping is obviously continuous and differentiable in both components, and surjectively maps the unit square to the unit segment.

@@

However this again fails because the joint cumulative function is not bijective, and isolines of the CDF define diagonals in the uniform embedding space $[0,1]^2$[^jointcdf].

[^jointcdf]: We note to avoid any confusion, that in @sec-mapping-cdf the CDF is taken component-wise, instead of jointly. The univariate gaussian CDF defines a bijection between $\R$ and $(0,1)$, but the joint CDF does not.

Perhaps the linear projection or CDF examples are a bit too naive, and we need to look for more complex examples. We might ask the question of whether a diffeomorphism (a differentiable bijection with differentiable inverse) even exists between the unit square and the unit segment. Unfortunately, results from topology give a negative answer, and even worse, it can be shown that there is no homeomorphism (a continuous bijection with continuous inverse) between the unit square and the unit segment. This is stated by Netto's theorem, which imposes continuous bijections to conserve the dimensionality between the domain and the image.

@@

Since in the composion $w_U \circ S^{-1}$ we work with the inverse mapping, it might be judicious to directly consider functions from $[0,1]$ into $[0,1]^2$. We have already established that there is no homeomorphism between those spaces, and need to relax one of the conditions.

@@

In the light of these results, we might be tempted to search for expressions for Cantor-type functions, since having a bijection would allow us to have a direct mapping between the square and the segment populations. Unfortunately, these functions are discontinuous which breaks the regularity of the kernel.

Peano functions seem like good candidates for the mapping $S^{-1}$, and are extensively studied in the litterature (we refer for instance to [@sagan_space-filling_1994;@Sagan1993]). They have the nice property that they are continuous, and are surjective, so we the segment populations are able to "reach" the entirety of the square populations. All things considered, non-injectivity is not a problem, since it just means that two segment populations can "refer" the same square population, which does not break the regularity of the kernel, since the two segments populations can just have their own neighbourhoods. Furthermore, we will see that in our examples, the non-injectivity can be ignored because the mapping is quasi-bijective.

@@

The intuition behind this expansion is we take a segment, we split it in two, write 0 to go left, write one to go right, then recurse. Then combine the bits of $v^{(n)}_1$ and $v^{(n)}_2$ to get the position in the $[0,1]^2$ embedding.

@@

In general, each bit $b^k$ of the image would be a binary function of the $n$ bits of $v_1$ and the $n$ bits of $v_2$, implicitly $b_k = b_k(b^1_1, b^1_2, \cdots, b^1_n, b^2_1, b^2_2, \cdots, b^2_n)$.
With this, we are able to formulate any mapping of $4^n$ distinct elements (here : squares of size $2^{-n} \times 2^{-n}$) onto $4^n$ elements (here : segments of length $4^{-n}$).

The intuition behind the use of $2n$ bits for the image is that we take $2^n \times 2^n = 4^n$ squares in $[0,1]^2$ and map those to $4^n$ segments in $[0,1]$. In an informal way, we map $n+n$ bits onto $2n$ bits.

@@

The constant functions are ignored, since they effectively just restrict the output domain to a subset of $[0,1]$ ; and since only $2n$ bits are allowed for the output, they are a "waste" by ignoring information of the input.
Because the don't add information to the output (they just "swap" the values of the output bits at known locations), inversion functions are also not considered.

<!-- [^inversion_restriction]: This last restriction is not strictly necessary for the following, however it helps to have a simple model to build a good understanding of the problem -->

@@

<!-- ```{dot}
digraph {
	subgraph {
		rank=same
		ordering="in"
		node [shape=square];
		edge [style=invis]           // for all edges, invisible links
		// we use the invisible edges to establish their sequence (kludge)
		cdots1 [label=⋯ shape=plaintext]
		cdots2 [label=⋯ shape=plaintext]
		b11 [label=b¹₁ shape=rect]
		b12 [label=b¹₂ shape=rect]
		b1n1 [label=b¹ₙ₋₁ shape=rect]
		b11 -> b12 -> cdots1 -> b1n1 -> b1n -> b21 -> b22 -> cdots2 -> b2n
	}

	subgraph {
		rank=same
		ordering="in"
		edge [style=invis]           // for all edges, invisible links
		// we use the invisible edges to establish their sequence (kludge)
		b1 -> b2 -> b3 -> b4 -> b5 -> b6 -> cdots0 -> b_2n1 -> b_2n
	}

	b11 -> b1
	b21 -> b2
	b12 -> b3
	b22 -> b4
	b1n -> b_2n1
	b2n -> b_2n
}
``` -->

@@

These simplifications might seem too restrictive, however we will show in the rest of this thesis that this formulation is flexible enough to express mapping which have good properties such as "locality", all while being simple enough to help build an intuition.

@@

With this formalism in mind, we give a geometric intrpretation to the mappings $S^n$. The core idea is that the populations in $[0,1]$ have a clear ordering to them, and we can enumerate them simply by their position in the embedding. Since we constructed $S^n$ to be bijective, this enumeration traces a path in the corresponding populations of the $[0,1]^2$ embedding.

@@

This way of enumerating the square populations might be familiar to people who have worked with numerical arrays. In such (contiguous) arrays, the values of are laid out in memory in "row-major" order (C-style) or "column-major" order (Fortran-style) ; and we can go from the 1D memory representation to a 2D matrix representation by using "reshape" operations. Effectively, this is how this mapping is implemented in code, using `numpy.ravel_multi_index` (2D to 1D) and `numpy.unravel_index` (1D to 2D).

@@

Research in computer science and efficient data structures has given rise to other ways of storing 2-dimensional information in a 1-dimensional memory.
A problem in GPU computing is that of texture locality : sampling of textures often involves reading data corresponding to a small 2D region, and therefore in order to improve speed (minimize cache misses), the 2D data should be packed close together in the 1D memory.
One way of ordering the 2D texture data in the 1D memory is by using the Z-order curve (also known as "Morton mapping" or "Lebesgue curve" @Sagan1994). In this way, points close in 2D *tend* to be tightly packed in 1D. We can define the Z-order curve (Z-mapping) in the following way[^zcurve] :

[^zcurve]: The Z-curve is often defined in the transpose manner, putting first the bits of $v_2$ (vertical component) then the bits of $v_1$, instead of $v_1$ then $v_2$. Doing this, we get the "upright" Z shapes, as [shown here](https://en.wikipedia.org/wiki/Z-order_curve#/media/File:Four-level_Z.svg).

@@

To illustrate this notion of locality, we take a few consecutive populations in 1D and show where the corresponding populations fall in the 2D embedding. This is done in @fig-locality, in which we see that the Z-mapping seems to conserve locality from 1D to 2D : populations close in 1D seem to be close in 2D.

We also see that the Column mapping is not local in this sense, since there is a large variation along the vertial direction, which seems to always be "of order one". This can be explained by looking at the expression for $C^{-1}_2(\alpha^{(n)}) = 0.b_{n+1} b_{n+2} \cdots b_{2n}$ : small variations of the order of $2^{-n}$ in the value of $\alpha$ always result in variations of order one in $C^{-1}_2$.

@@

The attentive reader might however object this treatment of locality, since in some sense the Column mapping is local, but from 2D to 1D. This can be seen from the definition of $\alpha^{(n)}=C(v^{(n)}_1,v^{(n)}_2) = 0.b^1_1 b^1_2 \cdots b^1_n b^2_1 b^2_2 \cdots b^2_n$, since small variations of order $2^{-n}$ in $v^{(n)}_1$ result in small variations also of order $2^{-n}$ in $\alpha^{(n)}$. Similarly, small variations of order $2^{-n}$ in $v^{(n)}_1$ result in even smaller variations also of order $2^{-2n}$ in $\alpha^{(n)}$ ! We therefore seem to have two notions of locality : from 1D to 2D, and from 2D to 1D ; and the Column mapping shows that they are not equivalent.

However, we argue that in our case only the notion of locality from 1D to 2D matters. In essence, the dynamics in the 2D embeddeding are already known, and we ask the question of whether it is possible to write an equivalent neural field 1D (in the sense that it has identical latent trajectories). Without repeating the intuition given in @sec-intro-simnf, if neighbouring populations in 1D have similar potentials, then this allows us to write a neural field equation in $[0,1]$. Therefore we would like to have the property that populations close in 1D are (inversely) mapped to populations close in 2D, which already have similar potentials.

@@

We now discuss what happens to the bijective mappings when $n\to\infty$. For this, let us consider again the "random mapping" considered in the introduction @sec-intro-simnf. We discussed how this hypothetical mapping maps each position in $[0,1]^2$ into a "random position" in $[0,1]$. A finite-$n$ formulation would be that the mapping corresponds to a *random permutation* of the bits. It is however unclear how to formally write this mapping in a way that it is well-defined for a given $n$, since by definition, it is random. For a given position of the input $(v_1,v_2)$, we would like that the image is stable in the $n\to\infty$ limit.

@@

This is the condition of "pointwise convergence" of the mappings. We might be tempted to write that any mapping for which we can write the binary expansion $S^n$ is pointwise convergent, however we show by the example of the "anti-Z" mapping that this is not the case. Let us define for finite $n$

@@

<!-- Anti-Z mapping. Also called in code "Far mapping" (initially I didn't see the link with the Z-order curve), because it destroys locality 2D -> 1D. -->

```{python}
#| label: fig-anti-z
#| fig-cap: Anti-Z mapping for $n=2$. Numbers represent the ordering of the populations in the $[0,1]$ embedding. We don't plot the trace, because it has many crossovers and overlaps and makes the plot hard to read.
#| echo: false
fig, ax = plot_trace(AntiZMapping, nrec=2, do_trace=False)
display(fig)
```

@fig-anti-z gives some insight into the problem of this mapping. For a fixed input $(v_1,v_2)$, we see that the output point position jumps around, and that these jumps do not decay, they are always of order 1. The intuition behind pointwise convergence is that as $n$ increases, the image of a point by $S^n$ is continually refined.

@@

In our setting, how can we therefore guarantee that the mappings are pointwise convergent ? We need to show that small fluctuations in the input result in small fluctuations of the output. The binary expansion allows us to argue that for the sequence of mappings to be pointwise convergent, we need that the "least significant bits" (LSB, the bits to the right of the binary expansion) of the input are mapped to the LSB of the output. Conversely, the "most significant bits" (MSB, the bits to the left of the binary expansion) should be mapped to the MSB of the output.

We illustrate this with the Z-mapping, written in @eq-z. Corrections of order $2^{-n}$ in $v_1$ and $v_2$ induce corrections of order $2^{1-2n}$ and $2^{-2n}$ in $\alpha$ respectively. Therefore, it is easy to see that the Z-mapping is pointwise convergent as $n\to\infty$. Similarly, the Column mapping in @eq-column is also pointwise convergent, since as discussed @sec-local it has locality of 2D to 1D.

#### Conservation of bijectivity in the $n\to\infty$ limit

However, the behavior of the Column mapping in that of the Z-mapping are very different in the $n\to\infty$ limit. Up until now, we have avoided this issue by considering the finite-$n$ approximation of the mappings, which by construction is bijective. We illustrate the infinite-$n$ (numerically, large $n$) limit of these two mappings in @fig-n-to-infty.

@@

We see that the Column mapping converges to projection on $v_1$, in other words $\alpha=C(v_1, v_2) = v_1$, showing that despite every $C^{n}$ being bijective, the limit of the sequence is not bijective. This becomes clear when looking at the binary expansion of $C$ :

@@

The Z-mapping however seems to converge to a fractal curve. Numerically of course, there seem to be infinite number of discontinuities, but the size of these discontinuities is overwhelmingly small, and *on average* the size of these numerical disconuities vanishes as $n\to\infty$. We don't go further into detail about this in this section (in @sec-zcurve-lebesgue we develop on the convergence to a space-filling fractal curve), but this gives some intuition for the arguments presented in @sec-locality.

We finish this discussion on the limit of the sequence of mappings by asking if there are other mappings similar to the Z-mapping, in the sense that they are pointwise convergent and the limit is bijective[^zmapping-notbijective]. In an informal way, we established two conditions for this in the context of the mappings described in @sec-mappings-binary :

[^zmapping-notbijective]: Despite what can often be found in online resources, the Z-mapping is *not* a bijection because that would violate Netto's theorem. The rew_ason for this is the ambiguity in the writing of binary expansions, and in @sec-zcurve-noninjective we give a proof that the Z-mapping is not injective in the $n\to\infty$ limit. However the Z-mapping is *quasi-bijective*, in the sense that the set of points for which the Z-mapping is not injective has zero measure, and can therefore be ignored.

1. condition of pointwise convergence : the LSB of the input must be mapped to the LSB of the output,
2. condition of "no information loss" : the MSB of the input must be mapped to the MSB of the output.

These two conditions do not leave a lot of flexibility in the design of new mappings. We can of course define mappings where we interleave the bits in a different order, for instance $0.b^2_1 b^1_1 b^2_2 b^1_2 \cdots$ ; or mappings where we reorder the $k$ bits of the input, for instance $0.b^1_2 b^2_2 b^1_1 b^2_1 b^1_4 b^2_4 b^1_3 b^2_3 \cdots$, but the conditions restrict the reordering to be permutations of neighbouring bits. In essence, these mappings are not fundamentally different, and the Z-mapping can be seen as a "canonical representation" of the pointwise convergent mappings with bijective limits.

@@

It is time for a little treat. Before we argue about how to simulate the dynamics of the 1D neural field, let us simulate the dynamics of the 2D neural field, then at each timestep, we map the 2D neural field to the 1D embedding. This will give us a reference of how the 1D neural field is supposed to look like under the condition that its latent dynamics are identical.

@@

how we numerically implemement a simulation on a fractal

In order to simulate the neural field numerically, we discretize by taking a grid of $4^n$ bins in the $[0,1]^2$ space.

$$
\int_{[0,1]^2} w_U(\vec v, \vec u) \phi(h_U(t, \vec u)) \mathrm d \vec u \rightarrow \sum_{j=1}^N \underbrace{w_U(\vec u_i, \vec u_j)}_{J_{ij}} \phi(\underbrace{h_U(t, \vec u_j)}_{h_j(t)})
$$

Each bin therefore corresponds to a population (density) indexed by $i = \{1\cdots4^n\}$. Population $i$ interacts with other populations through the (numerically finite) connectivity $J_{ij}$.

To simulate the neural field in $[0,1]$, we use the mapping $S(\vec u_i) \stackrel{\text{notation}}= S(i) = \alpha_i$ to embed the populations in 1D.

One trivial mapping would be $S(i) = i / 4^n$, which is just ordering the populations in [0,1] in the same order they appear inside the bins. More elaborate mappings are simply *permutations* of the numerically binned populations set $\{1\cdots4^n\}$, and the resulting connectivity matrix is then given by $J_{S(i),S(j)}$ [^1].

However, the connectivity matrix is invariant to permutations of neurons, and without further modification, the dynamics emerging from $J_{S(i),S(j)}$ are the same as the ones emerging from $J_{ij}$.

In the same way that discretizing the smooth 2D field works *because of* the smoothness, we discretize the 1D field emerging from the mapping. If the mapping has "nice properties" (e.g. "locality"), this discretization should work in a similar manner.

>>>>

## Population dynamics: an introductory example of a 1D embedding

@@

Let us for a moment consider another type of model for networks of neurons, that of homogeneous population dynamics, that describes how $N_\text{pop}$ different populations interact with each other. These types of models emerge when networks of neurons exhibit a finite number of "neuron types". We then group similar neurons ad make the approximation that all neurons inside the population have the same potential, and the same interaction weights with other neurons. In effect, every population then represents in actuality an infinity of discrete identical neurons.

@@

We don't show the derivation of population dynamics here (see @GerKis14, chapter 12.3 for a detailed analysis), but we point out the implicit scaling behavior of $\tilde J_{ab}$. Letting $i$ be a neuron of population $a$ and $j$ be a neuron of population $b$, the connectivity between the two neurons is $J_{ij}$ and the population connectivity scales with the number of neurons in the presynaptic population $|b|$. Furthermore, we make the reasonable assumption that $J_{ij}$ scales as $J^0_{ij} / N$, where $J^0_{ij} \sim \bO(1)$.

@@

So far, we have not considered that the populations are embedded in some abstract space; but the notion of population weights gives rise to the obvious embedding in $[0,1]$ given by the cumulative sums of the population weights, illustrated in @fig-popdynamics-01. We define (somewhat informally) the connectivity kernel $w(\alpha, \beta) = J^0_{ab}$ for $\alpha, \beta$ in the regions corresponding to populations $a$ and $b$ respectively.[^poplocation]

@@

When taken in isolation, the previous developments might sound a little bit silly, though they serve to give an impression of the problem at hand, and an interpretation of the rest of this work. For a finite number of populations $N_\text{pop}$, the population dynamics is perfectly defined, however what happens as $N_\text{pop} \to \infty$ ? The core of the question lies in the fact that we can easily *enumerate* a finite number of populations, but the sum $\sum_{b=1}^{N_\text{pop}}$ must have a clearly defined interpretation when this number tends to infinity.

In light of the toy model presented in @sec-nf, we can reinterpret the density $\rho$ of the neural field equation. Every point $\vec z \in \R^p$ (let us consider $p=2$ for the rest) describes an infinite population of homogeneous neurons, associated with a probability (density) $\rho(\vec z)$. Therefore, in order to write an equivalent neural field in $[0,1]$, we need a way to *enumerate* the populations. This is done using a mapping $S : [0,1]^2 \mapsto [0,1]$, where we traverse the $[0,1]^2$ space by following the order in the image $[0,1]$.

@@

For a finite number of populations $N_\text{pop}$, applying a bijective mapping $S$ will evidently give an equivalent neural field, because the connectivity matrix is permutation-invariant, and the resulting connectivity matrix $J_{S(a),S(b)}$ will therefore give equivalent dynamics. As we increase $N_\text{pop}$ by taking finer and finer grids, the image of the grids defines a sequence of mappings $S^n: \text{grid in}\ [0,1]^2 \mapsto \text{grid in}\ [0,1]$, which we aim to study in the rest of this chapter.

@@

The connectivity kernel $\tilde w$ defines the interactions between the populations in the $[0,1]$ embedding, which we define relating to the uniform kernel on $[0,1]^2$ as :

@@

When we do the change of variables with the CDF, we don't encounter problems because the Gaussian (inverse) CDF is differentiable and its derivative is continuous[^inversecdf]; therefore $w_U(\vec v, \vec u) = w(\textrm{CDF}^{-1}(\vec v), \textrm{CDF}^{-1}(\vec u))$ is still highly regular.

[^inversecdf]: Strictly speaking, we have the problem that $\textrm{CDF}^{-1}(v) \xrightarrow{v \to 1} \infty$, and so the derivative diverges when $v \to 1$ (and similarly, when $v \to 0$). However, this is not a problem, since these limits correspond to $z \to \pm \infty$, where the density vanishes.

@@

We start our search of mappings by attempting to find a mapping $S$ that is differentiable.
The easiest example of such a mapping would be a linear projection, for instance, $P(v_1, v_2) = v_1$. This mapping is obviously continuous and differentiable in both components, and surjectively maps the unit square to the unit segment.

@@

However, this again fails because the joint cumulative function is not bijective, and isolines of the CDF define diagonals in the uniform embedding space $[0,1]^2$.[^jointcdf]

[^jointcdf]: We note to avoid any confusion, that in @sec-mapping-cdf the CDF is taken component-wise, instead of jointly. The univariate Gaussian CDF defines a bijection between $\R$ and $(0,1)$, but the joint CDF does not.

Perhaps the linear projection or CDF examples are a bit too naive, and we need to look for more complex examples. We might ask the question of whether a diffeomorphism (a differentiable bijection with a differentiable inverse) even exists between the unit square and the unit segment. Unfortunately, results from topology give a negative answer, and even worse, it can be shown that there is no homeomorphism (a continuous bijection with continuous inverse) between the unit square and the unit segment. This is stated by Netto's theorem, which imposes continuous bijections to conserve the dimensionality between the domain and the image.

@@

Since in the composition $w_U \circ S^{-1}$ we work with the inverse mapping, it might be judicious to directly consider functions from $[0,1]$ into $[0,1]^2$. We have already established that there is no homeomorphism between those spaces, and need to relax one of the conditions.

@@

In light of these results, we might be tempted to search for expressions for Cantor-type functions, since having a bijection would allow us to have a direct mapping between the square and the segment populations. Unfortunately, these functions are discontinuous which breaks the regularity of the kernel.

Peano functions seem like good candidates for the mapping $S^{-1}$, and are extensively studied in the literature (we refer for instance to [@sagan_space-filling_1994;@Sagan1993]). They have the nice property that they are continuous and surjective, so we the segment populations can "reach" the entirety of the square populations. All things considered, non-injectivity is not a problem, since it just means that two segment populations can "refer" to the same square population, which does not break the regularity of the kernel, since the two segment populations can just have their own neighborhoods. Furthermore, we will see that in our examples, the non-injectivity can be ignored because the mapping is quasi-bijective.

@@

The intuition behind this expansion is we take a segment, split it in two, write 0 to go left, write one to go right, then recurse. Then combine the bits of $v^{(n)}_1$ and $v^{(n)}_2$ to get the position in the $[0,1]^2$ embedding.

@@

In general, each bit $b^k$ of the image would be a binary function of the $n$ bits of $v_1$ and the $n$ bits of $v_2$, implicitly $b_k = b_k(b^1_1, b^1_2, \cdots, b^1_n, b^2_1, b^2_2, \cdots, b^2_n)$.
With this, we can formulate any mapping of $4^n$ distinct elements (here : squares of size $2^{-n} \times 2^{-n}$) onto $4^n$ elements (here : segments of length $4^{-n}$).

The intuition behind the use of $2n$ bits for the image is that we take $2^n \times 2^n = 4^n$ squares in $[0,1]^2$ and map those to $4^n$ segments in $[0,1]$. Informally, we map $n+n$ bits onto $2n$ bits.

@@

The constant functions are ignored, since they effectively just restrict the output domain to a subset of $[0,1]$; and since only $2n$ bits are allowed for the output, they are a "waste" by ignoring information of the input.
Because they don't add information to the output (they just "swap" the values of the output bits at known locations), inversion functions are also not considered.

@@

@@

These simplifications might seem too restrictive, however, we will show in the rest of this thesis that this formulation is flexible enough to express mappings that have good properties such as "locality", all while being simple enough to help build intuitions.

@@

With this formalism in mind, we give a geometric interpretation to the mappings $S^n$. The core idea is that the populations in $[0,1]$ have a clear ordering to them, and we can enumerate them simply by their position in the embedding. Since we constructed $S^n$ to be bijective, this enumeration traces a path in the corresponding populations of the $[0,1]^2$ embedding.

@@

This way of enumerating the square populations might be familiar to people who have worked with numerical arrays. In such (contiguous) arrays, the values are laid out in memory in "row-major" order (C-style) or "column-major" order (Fortran-style); and we can go from the 1D memory representation to a 2D matrix representation by using "reshape" operations. Effectively, this is how this mapping is implemented in code, using `numpy.ravel_multi_index` (2D to 1D) and `numpy.unravel_index` (1D to 2D).

@@

Research in computer science and efficient data structures has given rise to other ways of storing 2-dimensional information in a 1-dimensional memory.
A problem in GPU computing is that of texture locality: sampling textures often involves reading data corresponding to a small 2D region, and therefore to improve speed (minimize cache misses), the 2D data should be packed close together in the 1D memory.
One way of ordering the 2D texture data in the 1D memory is by using the Z-order curve (also known as "Morton mapping" or "Lebesgue curve" @Sagan1994). In this way, points close in 2D *tend* to be tightly packed in 1D. We can define the Z-order curve (Z-mapping) in the following way[^zcurve] :

[^zcurve]: The Z-curve is often defined in a transpose manner, putting first the bits of $v_2$ (vertical component) then the bits of $v_1$, instead of $v_1$ then $v_2$. By doing this, we get the "upright" Z shapes, as [shown here](https://en.wikipedia.org/wiki/Z-order_curve#/media/File:Four-level_Z.svg).

@@

To illustrate this notion of locality, we take a few consecutive populations in 1D and show where the corresponding populations fall in the 2D embedding. This is done in @fig-locality, in which we see that the Z-mapping seems to conserve locality from 1D to 2D: populations close in 1D seem to be close in 2D.

We also see that the Column mapping is not local in this sense, since there is a large variation along the vertical direction, which seems to always be "of order one". This can be explained by looking at the expression for $C^{-1}_2(\alpha^{(n)}) = 0.b_{n+1} b_{n+2} \cdots b_{2n}$ : small variations of the order of $2^{-n}$ in the value of $\alpha$ always result in variations of order one in $C^{-1}_2$.

@@

The attentive reader might however object to this treatment of locality, since in some sense the Column mapping is local, but from 2D to 1D. This can be seen from the definition of $\alpha^{(n)}=C(v^{(n)}_1,v^{(n)}_2) = 0.b^1_1 b^1_2 \cdots b^1_n b^2_1 b^2_2 \cdots b^2_n$, since small variations of order $2^{-n}$ in $v^{(n)}_1$ result in small variations also of order $2^{-n}$ in $\alpha^{(n)}$. Similarly, small variations of order $2^{-n}$ in $v^{(n)}_1$ result in even smaller variations also of order $2^{-2n}$ in $\alpha^{(n)}$ ! We therefore seem to have two notions of locality: from 1D to 2D, and from 2D to 1D; and the Column mapping shows that they are not equivalent.

However, we argue that in our case only the notion of locality from 1D to 2D matters. In essence, the dynamics in the 2D embedding are already known, and we ask the question of whether it is possible to write an equivalent neural field 1D (in the sense that it has identical latent trajectories). Without repeating the intuition given in @sec-intro-simnf, if neighboring populations in 1D have similar potentials, then this allows us to write a neural field equation in $[0,1]$. Therefore we would like to have the property that populations close in 1D are (inversely) mapped to populations close in 2D, which already have similar potentials.

@@

We now discuss what happens to the bijective mappings when $n\to\infty$. For this, let us consider again the "random mapping" considered in the introduction @sec-intro-simnf. We discussed how this hypothetical mapping maps each position in $[0,1]^2$ into a "random position" in $[0,1]$. A finite-$n$ formulation would be that the mapping corresponds to a *random permutation* of the bits. It is unclear how to formally write this mapping in a way that is well-defined for a given $n$, since by definition, it is random. For a given position of the input $(v_1,v_2)$, we would like that the image is stable in the $n\to\infty$ limit.

@@

This is the condition of "pointwise convergence" of the mappings. We might be tempted to write that any mapping for which we can write the binary expansion $S^n$ is pointwise convergent, however, we show by the example of the "anti-Z" mapping that this is not the case. Let us define for finite $n$

@@

```{python}
#| label: fig-anti-z
#| fig-cap: Anti-Z mapping for $n=2$. Numbers represent the ordering of the populations in the $[0,1]$ embedding. We don't plot the trace, because it has many crossovers and overlaps which makes the plot hard to read.
#| echo: false
fig, ax = plot_trace(AntiZMapping, nrec=2, do_trace=False)
display(fig)
```

@fig-anti-z gives some insight into the problem of this mapping. For a fixed input $(v_1,v_2)$, we see that the output point position jumps around and that these jumps do not decay, they are always of order 1. The intuition behind pointwise convergence is that as $n$ increases, the image of a point by $S^n$ is continually refined.

@@

In our setting, how can we guarantee that the mappings are pointwise convergent? We need to show that small fluctuations in the input result in small fluctuations in the output. The binary expansion allows us to argue that for the sequence of mappings to be pointwise convergent, we need that the "least significant bits" (LSB, the bits to the right of the binary expansion) of the input are mapped to the LSB of the output. Conversely, the "most significant bits" (MSB, the bits to the left of the binary expansion) should be mapped to the MSB of the output.

We illustrate this with the Z-mapping, written in @eq-z. Corrections of order $2^{-n}$ in $v_1$ and $v_2$ induce corrections of order $2^{1-2n}$ and $2^{-2n}$ in $\alpha$ respectively. Therefore, it is easy to see that the Z-mapping is pointwise convergent as $n\to\infty$. Similarly, the Column mapping in @eq-column is also pointwise convergent, since as discussed @sec-local it has a locality of 2D to 1D.

#### Conservation of bijectivity in the $n\to\infty$ limit {#sec-ninfty-bijection}

However, the behavior of the Column mapping and that of the Z-mapping are very different in the $n\to\infty$ limit. Up until now, we have avoided this issue by considering the finite-$n$ approximation of the mappings, which by construction is bijective. We illustrate the infinite-$n$ (numerically, large $n$) limit of these two mappings in @fig-n-to-infty.

@@

We see that the Column mapping converges to projection on $v_1$, in other words, $\alpha=C(v_1, v_2) = v_1$, showing that despite every $C^{n}$ being bijective, the limit of the sequence is not bijective. This becomes clear when looking at the binary expansion of $C$ :

@@

The Z-mapping however seems to converge to a fractal curve. Numerically, there seems to be an infinite number of discontinuities, but the size of these discontinuities is overwhelmingly small, and *on average* the size of these numerical discontinuities vanishes as $n\to\infty$. We won't go further into detail about this in this section (in @sec-zcurve-lebesgue we develop the convergence to a space-filling fractal curve), but this gives some intuition for the arguments presented in @sec-locality.

We finish this discussion on the limit of the sequence of mappings by asking if there are other mappings similar to the Z-mapping, in the sense that they are pointwise convergent and the limit is bijective.[^zmapping-notbijective] In an informal way, we established two conditions for this in the context of the mappings described in @sec-mappings-binary :

[^zmapping-notbijective]: Despite what can often be found in online resources, the Z-mapping is *not* a bijection because that would violate Netto's theorem. The reason for this is the ambiguity in the writing of binary expansions, and in @sec-zcurve-noninjective we give proof that the Z-mapping is not injective in the $n\to\infty$ limit. However, the Z-mapping is *quasi-bijective*, in the sense that the set of points for which the Z-mapping is not injective has zero measure, and can therefore be ignored.

1. condition of pointwise convergence: the LSB of the input must be mapped to the LSB of the output,
2. condition of "no information loss": the MSB of the input must be mapped to the MSB of the output.

These two conditions do not leave a lot of flexibility in the design of new mappings. We can evidently define mappings where we interleave the bits in a different order, such as $0.b^2_1 b^1_1 b^2_2 b^1_2 \cdots$; or mappings where we reorder the $k$ bits of the input, as in $0.b^1_2 b^2_2 b^1_1 b^2_1 b^1_4 b^2_4 b^1_3 b^2_3 \cdots$, but the conditions restrict the reordering to be permutations of neighboring bits. In essence, these mappings are not fundamentally different, and the Z-mapping can be seen as a "canonical representation" of the pointwise convergent mappings with bijective limits.

@@

It is time for a little treat. Before we argue about how to simulate the dynamics of the 1D neural field, let us simulate the dynamics of the 2D neural field, then at each timestep, we map the 2D neural field to the 1D embedding. This will give us a reference of how the 1D neural field is supposed to look under the condition that its latent dynamics are identical.

@@

### Numerical aspects of implementing the mappings

In the previous sections, we have discussed what mappings have good properties, and have given a few hints as to how we can implement them numerically. Since the mappings we presented can be described as a binary expansion, they naturally lend themselves to numerical approximations. Furthermore, this is made especially easy since the mappings we defined simply are permutations of bits (in the $n < \infty$ case), and so we can write a one-to-one mapping between (unsigned integers).

Without going into too much detail on the programming, we have implemented the mappings as all inheriting from the same abstract class, and each mapping simply has to define a method `indices_to_indices2d(indices: array[int, 1]) -> array[int, 2]` to go from 1D to 2D ($S^{-1}$), and a method `indices2d_to_indices(indices2d: array[int, 2]) -> array[int, 1]` to go from 2D to 1D.[^indices] The abstract class simply defines additional quality-of-life functionality, which makes use of these two methods.

[^indices]: In the source code, only the method `indices(v12: array[float, 2]) -> array[int, 1]` is defined. This is because some mappings are implemented (for instance, the `RecursiveLocalMapping`, that we don't discuss here), which don't lend themselves to the simple permutation of bits formulation.

### Numerical approximations of a fractal structure

In the continuation of the simulation methods established in the @sec-simnf, we adopt a grid discretization scheme, as it naturally lends itself to be applied with the mappings. We however note that using binning methods, the Monte-Carlo simulation can still be applied. The intuition behind this is that sampling populations inside a small square of size $2^{-n} \times 2^{-n}$ gives rise to the notion of an "average connectivity kernel", and in our case the notion of "mean patterns". More details are given in @sec-mean-patterns.
